Algorithms for Estimating Trends in Global Temperature Volatility

### 2018-11-07 aaai-2018/khodadadi-mcdonald-aaai19-camera.tex
update ignore

\pdfinfo{
  /Title (Algorithms for Estimating Trends in Global Temperature Volatility)
   %/Author (Khodadadi and McDonald)
  /Author (John Doe, Jane Doe)}

@@

\newcommand{\citet}[1]
{\citeauthor{#1} ̃\shortcite{#1}}
\newcommand{\citep}{\cite}
\newcommand{\citealp}[1]
{\citeauthor{#1} ̃\citeyear{#1}}

@@

%\newcommand{\theHalgorithm}{\arabic{algorithm}}
\usepackage{cleveref}

@@

@@

\author{Author 1 \ and Author 2\\
Address line\\
Address line
}
% \author{Arash Khodadadi \and Daniel J. McDonald\\
%  Department of Statistics\\
%  Indiana University\\
%  Bloomington, IN 47408 \\
%  \{arakhoda,dajmcdon\}@indiana.edu}

@@

Our main contributions are as follows:
\begin{enumerate}
\item We propose a method for nonparametric variance estimation for a
  spatio-temporal process and discuss the relationship between our
  methods and those existing in the machine learning literature (\autoref{sec:ell_1-trend-filt}).
\item We derive two alternating direction method of multiplier
  algorithms to fit our estimator when applied to 
  very large data (\autoref{sec:prop-optim-meth}). We give situations
  under which each algorithm is most likely to be useful. Open-source
  Python code is available on {\tt github}.%\footnote{Link removed for anonymity.}
\item Because the construction of satellite-based datasets is ongoing
  and currently proprietary, we illustrate our methods on a large,
  publicly available, global temperature dataset.
  The goal is to demonstrate the feasibility of these methods for tracking
  world-wide trends in variance in standard atmospheric data and a
  simulation constructed to mimic these data's features
  (\autoref{sec:empirical-evaluation}). 
\end{enumerate}

While the motivation for our methodology is its application to large, gridded climate
data, we note that our algorithms are also applicable to neuroimaging,
image denoising, or examining large collections of financial instruments.

@@

We estimated the linearized ADMM for all combinations of values of
$\lambda_t$ and $\lambda_s$ from the sets $\lambda_t \in
\{0,1,5,10,50,100\}$ and $\lambda_s \in \{0,0.05,0.1,0.2,0.3\}$. For
each pair, we then compute the mean absolute error (MAE) between the
estimated variance and the true variance at all locations and all time
steps. For $\lambda_t=5$ and $\lambda_s=0.1$, the MAE was minimized. 
\begin{figure}[tb]
  \centering	
  \includegraphics[width=.48\columnwidth]{Figures/convergence}  
  \includegraphics[width=.48\columnwidth]{Figures/modelComp_MAE}  
  \caption{Left: Value of the objective function for linearized (orange)
    and consensus (blue) ADMM against iteration. Right: MAE for (1) our
    method with optimal values of $\lambda_t$ and $\lambda_s$ (2)
    spatial penalty only (3) temporal
    penalty only and (4) a GARCH(1,1).} 
  \label{fig:true_fitted_var}
\end{figure}
The
bottom panel of \autoref{fig:true_var_spatial} shows the true and the
estimated standard deviation at location (0,0) 
and $\lambda_t=5$ (blue) and $\lambda_t=100$ (green) ($\lambda_s=0.1$).
Larger values of $\lambda_t$ lead to estimated values
which are ``too smooth". The left panel of \autoref{fig:true_fitted_var} shows the
convergence of both Algorithms as a function of iteration. It is
important to note that each iteration of the linearized
algorithm takes 0.01 seconds on average while each iteration of the
consensus ADMM takes about 20 seconds. Thus, where the lines meet at
400 iterations requires about 4 seconds for the linearized method and
2 hours for the consensus method.

@@

As shown in \autoref{alg:linADMM}, we checked convergence using
$\epsilon=0.001\%$ of the MSE of the data.
Our simulations indicated that the
convergence speed depends on the value of $\lambda_t$ and
$\lambda_s$. For the temperature data, we used the solutions obtained for smaller values
of these parameters as a warm start for the larger values resulting in
faster convergence.

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=.66\columnwidth]{Figures/ts_estimatedVarLeft}
  \includegraphics[width=.66\columnwidth]{Figures/ts_estimatedVarRight}
  \caption{Residuals from the estimated trend
    (blue), the estimated SDs (orange), and annual average SD (green) for
   Indianapolis (top) and San Diego (bottom). Units are $\textrm{K}^{\circ}$.} 
  \label{fig:avg_change_estimatedSD}
\end{figure}

\autoref{fig:avg_change_estimatedSD} shows the
detrended data, the estimated standard deviation and the yearly
average of these estimates for two cities in the US: Indianapolis
(top) and San Diego (bottom). The estimated SD 
captures the periodic behavior in the variance of the time-series. In
addition, the number of linear segments changes adaptively in each
time window depending on how fast the variance is changing.  
\begin{figure}[tb]
  \centering
  \includegraphics[width=.9\linewidth]{Figures/avg_logVar.pdf}\\
  \includegraphics[width=.9\linewidth]{Figures/avg_change_logVar.pdf}
  \caption{The
    average of the detrended estimated variance over the northern hemisphere (top) and the
    change in the variance from 1961 to 2011 (bottom). Units are K$^\circ$.}
\label{fig:avg_chg}
\end{figure}

@@

>>>>

\pdfinfo{
  /Title (Algorithms for Estimating Trends in Global Temperature Volatility)
   /Author (Khodadadi and McDonald)
   % /Author (John Doe, Jane Doe)
 }

@@

\newcommand{\citet}[1]
{\citeauthor{#1} ~\shortcite{#1}}
\newcommand{\citep}{\cite}
\newcommand{\citealp}[1]
{\citeauthor{#1} ~\citeyear{#1}}

@@

%\newcommand{\theHalgorithm}{\arabic{algorithm}

@@

\usepackage{cleveref}

@@

% \author{Author 1 \ and Author 2\\
% Address line\\
% Address line
% }
\author{Arash Khodadadi \and Daniel J. McDonald\\
 Department of Statistics\\
 Indiana University\\
 Bloomington, IN 47408 \\
 \{arakhoda,dajmcdon\}@indiana.edu}

@@

Our main contributions are as follows:
\begin{enumerate}
\item We propose a method for nonparametric variance estimation for a
  spatio-temporal process and discuss the relationship between our
  methods and those existing in the machine learning literature (\autoref{sec:ell_1-trend-filt}).
\item We derive two alternating direction method of multiplier
  algorithms to fit our estimator when applied to 
  very large data (\autoref{sec:prop-optim-meth}). We give situations
  under which each algorithm is most likely to be useful. Open-source
  Python code is available.\footnote{\tt github.com/dajmcdon/VolatilityTrend}
\item Because the construction of satellite-based datasets is ongoing
  and currently proprietary, we illustrate our methods on a large,
  publicly available, global temperature dataset.
  The goal is to demonstrate the feasibility of these methods for tracking
  world-wide trends in variance in standard atmospheric data and a
  simulation constructed to mimic these data's features
  (\autoref{sec:empirical-evaluation}). 
\end{enumerate}

While the motivation for our methodology is its application to large, gridded climate
data, we note that our algorithms are easily generalizable to
spatio-temporal data under convex loss, e.g. exponential family
likelihood. Furthermore the spatial structure can be broadly construed
to include general graph dependencies. Our current application uses
Gamma likelihood which lends itself well to modeling trends in
pollutant emissions or in astronomical phenomena like microwave
background radiation. Volatility estimation in natural gas/oil markets
or with financial data is another possibility. These methods can also
be applied to resting-state fMRI data (though the penalty structure changes).

@@

We estimated the linearized ADMM for all combinations of values of
$\lambda_t$ and $\lambda_s$ from the sets $\lambda_t \in
\{0,1,5,10,50,100\}$ and $\lambda_s \in \{0,0.05,0.1,0.2,0.3\}$. For
each pair, we then compute the mean absolute error (MAE) between the
estimated variance and the true variance at all locations and all time
steps. For $\lambda_t=5$ and $\lambda_s=0.1$, the MAE was minimized. 
\begin{figure}[tb]
  \centering	
  \includegraphics[width=.48\columnwidth]{Figures/convergence}  
  \includegraphics[width=.48\columnwidth]{Figures/modelComp_MAE}  
  \caption{Left: Value of the objective function for linearized (orange)
    and consensus (blue) ADMM against iteration. Right: MAE for (1) our
    method with optimal values of $\lambda_t$ and $\lambda_s$ (2)
    spatial penalty only (3) temporal
    penalty only and (4) a GARCH(1,1).} 
  \label{fig:true_fitted_var}
\end{figure}
The
bottom panel of \autoref{fig:true_var_spatial} shows the true and the
estimated standard deviation at location (0,0) 
and $\lambda_t=5$ (blue) and $\lambda_t=100$ (green) ($\lambda_s=0.1$).
Larger values of $\lambda_t$ lead to estimated values
which are ``too smooth". The left panel of \autoref{fig:true_fitted_var} shows the
convergence of both Algorithms as a function of iteration. It is
important to note that each iteration of the linearized
algorithm takes 0.01 seconds on average while each iteration of the
consensus ADMM takes about 20 seconds. Thus, where the lines meet at
400 iterations requires about 4 seconds for the linearized method and
2 hours for the consensus method. For consensus ADMM, computation per
iteration per core is requires $\sim$10 seconds with $\sim$4 seconds
for communication. In general, the literature suggests linear
convergence for ADMM~\citep{NishiharaLessard2015} and~\autoref{fig:true_fitted_var} seems to fit in the linear framework for both algorithms, though with different constants.

@@

As shown in \autoref{alg:linADMM}, we checked convergence using
$\epsilon=0.001\%$ of the MSE of the data.
Our simulations indicated that the
convergence speed depends on the value of $\lambda_t$ and
$\lambda_s$. For the temperature data, we used the solutions obtained for smaller values
of these parameters as a warm start for the larger values resulting in
faster convergence. The timing of the estimation depends on $\lambda_t$ and $\lambda_s$ but takes between 1 and 4 hours for convergence.

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=.45\columnwidth]{Figures/ts_estimatedVarLeft}
  \includegraphics[width=.45\columnwidth]{Figures/ts_estimatedVarRight}
  \caption{Residuals from the estimated trend
    (blue), the estimated SDs (orange), and annual average SD (green) for
   Indianapolis (left) and San Diego (right). Units are $\textrm{K}^{\circ}$.} 
  \label{fig:avg_change_estimatedSD}
\end{figure}

\autoref{fig:avg_change_estimatedSD} shows the
detrended data, the estimated standard deviation and the yearly
average of these estimates for two cities in the US: Indianapolis
(left) and San Diego (right). The estimated SD 
captures the periodic behavior in the variance of the time-series. In
addition, the number of linear segments changes adaptively in each
time window depending on how fast the variance is changing.  
\begin{figure}[tb]
  \centering
  \includegraphics[width=.9\linewidth]{Figures/avg_logVar.pdf}\\
  \includegraphics[width=.9\linewidth]{Figures/avg_change_logVar.pdf}
  \caption{The
    average of the detrended estimated variance over the northern hemisphere (top) and the
    change in the variance from 1961 to 2011 (bottom). Units are K$^\circ$.}
\label{fig:avg_chg}
\end{figure}

@@

\section*{Acknowledgements}

This material is based upon work supported by the National Science
Foundation under Grant Nos.\ DMS--1407439 and DMS--1753171.
